The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-08 manuscript/manuscript.Rnw
comment out comparative analysis of linage

We performed a comparative analysis of gene expression by
reconstructing the evolution of expression values and detecting
significant shifts in expression along specific lineages. Likelihood
values for each model for each of the X genes reconstructed, as well
as likelihood ratio tests, P- and Q-values (after correcting for
multiple testing) are shown in Supplemental Table X. We identified
between 4 (Boreoeutheria) and 550 (Mouse) significant shifts (Q <
0.01) in expression on specific lineages (Table 10). The top 10 genes
with significant shifts (if applicable) per lineage are shown in Table
10 and all genes for each branch are shown in Supplemental Tables
X-XX. Genes that showed shifts on the human lineage were
overwhelmingly identified as placenta-specific by the
tissue-specificity index (TSI), with 56 out of 279 genes (20\%) with
both TSI>0.75 and placenta as one the top two tissues with the highest
expression. Unsurpisingly, this was not the case with other lineages,
since we used expression values from a human-derived microarray to
generate TSI values for each gene. However, this does underscore the
dynamic nature of expression patterns in the placenta and indeed all
tissues across mammals.

We then performed DAVID functional annotation clustering analyses on
each list of genes from each lineage. The top three clusters for each
lineage are shown in Table 10 and a full list of clusters in
Supplemental Table X. The top enrichment cluster in the human lineage
is blood vessel development, followed by extracellular matrix and
metal ion binding (Table 10). Other significant clusters include those
related to cell communication/signaling, cytokines, defense response,
tryptophan metabolism, apoptosis, and skeletal system development
(Supplemental Table X). The bonobo lineage was enriched in only three
clusters: apoptosis, Toll-like receptor signalling and apoptosis. The
mouse displayed enrichment in clusters involved in epigenetic
modifications such as methyltransferase activity and histone
modification, as well embryonic morphogeneisis, lipid biosynthesis,
among others (Supplemental Table X). The top enriched cluster in the
cow lineage was the response to vitamin A, as well as many clusters
related to the immune system. The dog showed enrichment in skeletal
system development, as well as lipid metabolic processes and blood
circulation. The elephant lineage had only one enriched cluster:
carbohydrate binding. Genes that showed enrichment on the opossum
lineage (essentially representing the difference between placental
mammals and marsupials) were clustered in many categories not seen in
the other lineages, including clusters related to the mitochondria,
urea cycle, peroxisome, glycolysis, and transport across the cell
membrane (Table 10 and Supplemental Table X).

>>>>

% Not sure if this is actually worthwhile; I've mentioned some of this
% already for some of the lineages

% \subsection{Comparative Analysis of Gene Expression}
% 
% We performed a comparative analysis of gene expression by
% reconstructing the evolution of expression values and detecting
% significant shifts in expression along specific lineages. Likelihood
% values for each model for each of the X genes reconstructed, as well
% as likelihood ratio tests, P- and Q-values (after correcting for
% multiple testing) are shown in Supplemental Table X. We identified
% between 4 (Boreoeutheria) and 550 (Mouse) significant shifts (Q <
% 0.01) in expression on specific lineages (Table 10). The top 10 genes
% with significant shifts (if applicable) per lineage are shown in Table
% 10 and all genes for each branch are shown in Supplemental Tables
% X-XX. Genes that showed shifts on the human lineage were
% overwhelmingly identified as placenta-specific by the
% tissue-specificity index (TSI), with 56 out of 279 genes (20\%) with
% both TSI>0.75 and placenta as one the top two tissues with the highest
% expression. Unsurpisingly, this was not the case with other lineages,
% since we used expression values from a human-derived microarray to
% generate TSI values for each gene. However, this does underscore the
% dynamic nature of expression patterns in the placenta and indeed all
% tissues across mammals.
% 
% We then performed DAVID functional annotation clustering analyses on
% each list of genes from each lineage. The top three clusters for each
% lineage are shown in Table 10 and a full list of clusters in
% Supplemental Table X. The top enrichment cluster in the human lineage
% is blood vessel development, followed by extracellular matrix and
% metal ion binding (Table 10). Other significant clusters include those
% related to cell communication/signaling, cytokines, defense response,
% tryptophan metabolism, apoptosis, and skeletal system development
% (Supplemental Table X). The bonobo lineage was enriched in only three
% clusters: apoptosis, Toll-like receptor signalling and apoptosis. The
% mouse displayed enrichment in clusters involved in epigenetic
% modifications such as methyltransferase activity and histone
% modification, as well embryonic morphogeneisis, lipid biosynthesis,
% among others (Supplemental Table X). The top enriched cluster in the
% cow lineage was the response to vitamin A, as well as many clusters
% related to the immune system. The dog showed enrichment in skeletal
% system development, as well as lipid metabolic processes and blood
% circulation. The elephant lineage had only one enriched cluster:
% carbohydrate binding. Genes that showed enrichment on the opossum
% lineage (essentially representing the difference between placental
% mammals and marsupials) were clustered in many categories not seen in
% the other lineages, including clusters related to the mitochondria,
% urea cycle, peroxisome, glycolysis, and transport across the cell
% membrane (Table 10 and Supplemental Table X).
